Investigating Determinism of Game Engines for Autonomous Vehicle Verification

### 2020-04-20 paper/cav_determinisim_main.tex
GC comments

% The use of game engines for AV testing
Increasingly, the autonomous vehicle (AV) community is adopting game engines as simulation platforms to support the development and testing of vehicle control software. 
%
Carla\footnote{http://carla.org/}, for instance, is an open-source simulator for autonomous driving that is implemented as a layer over the Unreal Engine (UE4), a  real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.

@@

% Game engines for SBV
Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a convenient simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation. However, several challenges remain; one of these is determinism.
%
% What is a deterministic system
Deterministic, in this context, refers to the property of causality given a temporal development of events such that any state is completely determined by prior states.
%
Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly under identical conditions. Specifically, we focus on the variance of actor paths within a given scenario.

We will adopt the terminology defined in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements. Dynamic elements are the elements in a scene whose actions or behaviour may change over time; these are considered actors and may include other road vehicles, cyclists, pedestrians and traffic signals. The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific goals and values.\todo[inline]{values - Does this mean stimulus or other parameters?} A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.

@@

% Challenges of SBV
Game engines do not need to be deterministic when used for gaming; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers. Because the situation is very different for AV development and testing, our main research question is: 
how can one assess the extent to which a simulation environment is deterministic? 
%
In pursuit of an answer to this question we have developed a method to determine the simulation variance for a given simulation environment. 
%
Knowing the simulation variance will help developers and testers understand the impact of using a game engine for AV simulation, especially how potentially existing non-determinism affects repeatability and reliability.\todo[inline]{review and potentially summarize results from case study}
%
%The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.\todo[inline]{next}

This paper is structured as follows. In Section~\ref{s:background} game engines and sources of potential non-determinism in game engines are investigated. Section~\ref{s:methodology} presents a method to assess the suitability of a simulation system for AV verification. 
%
A case study to illustrate how the simulation variance observed for a number of scenarios is presented in Section~\ref{s:case-study}.\todo[inline]{point to UE4} We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo[inline]{review at end}

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoop.pdf}
\caption{Game engine loop block diagram~\cite{GameProgPatternsBook}}
\label{GameEngineLoopDiagram}
\end{figure}

The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of AV development, the response of the AV to the current state of its environment. For example, the throttle and steering angle from the controller of an AV. The update stage interprets the input from all dynamic actors in the scene, performs any necessary physics calculations and updates the actor states. The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}. \todo[inline]{Fig 1: update rates of which clock?}\todo[inline]{Should there not also be a clock in the overall loop?}\todo[inline]{Greg to look up Unity wrt concurrently running processes and clocks.}

@@

As the game progresses, the game loop runs continuously according to the game clock.\todo[inline]{introduce the different clocks and what they govern, real world, game, physics and render}
%
At the beginning of each game clock tick, the lag\todo[inline]{first explain where the lag comes from} between the game clock and the real world is updated based on how much real time has passed. This measures how far the game's clock is behind compared to the real world. 
%
Then the user inputs are processed.
%
There is then an inner loop to update the physics processes (clock symbol in Fig.~\ref{GameEngineLoopDiagram}), incrementing at a delta time step, $dt$, until the game clock is equal to the real world. 
%
Rendering occurs once the game clock has caught up with the real world clock. 
%
The process then repeats.

The physics engine operates according to a time step, $dt$. %, because it makes everything simpler and more stable for physics and AI.
The shorter this time step is, the smoother the interpretation of the physical dynamics will be. %processing time it takes to catch up to game clock. % and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}. 
%
Fixing the physics time step to a constant $dt$ results in simpler and more stable physics simulation and AI. 
%
To use a fixed physics time step the user's display refresh rate needs to be known in advance. This requires an update loop to take less than one render tick (one frame of real world time). Given the range of different hardware capabilities, a variable delta time is often implemented for game playing, taking the previous frame time as the next $dt$. However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations~\cite{gaffer}. 
%
Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allows computational headroom for slower hardware. Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/}}. 
%
If the engine tries to render between physics updates, \textit{residual lag} can occur and extrapolation between frames needs to be performed to smooth transition between scenes. This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}. \cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited?}\todo[inline]{Why is this part of the figure? It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other.}
%
% Ideally the time step should be as short as possible, so that the simulation returns better accuracy of complex physical processes. %runs with high fidelity on fast machines. 
In exceptional cases where resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines, 
and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour.\todo[inline]{Wrt the figure, is ``lag'' not a delay, i.e.\ should happen after the tick, rather than before?}

@@

In the context of AV simulation, this could result in precision issues of, for example, actor position within the environment leading to falsely satisfied or failed assertions.\todo[inline]{state what assertions are (used for)} 

@@

%Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}. 

@@

\\\\
\noindent\underline{\textit{Error Correcting Code (ECC) Memory:}}
ECC Memory can detect and correct single bit errors in DRAM memory. Single bit errors may occur due to malfunctioning hardware, ionising radiation (background cosmic or environmental sources) or from electromagnetic radiation~\cite{dodd2003basic}. If single bit errors go uncorrected then subsequent computational processing will produce incorrect results, potentially giving rise to non-determinism due to the probabilistic nature of such errors occurring. Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.\todo[inline]{that said, HW simulation is deterministic, uses long simulation runs and thousands of CPU hours - I've not heard that this is an issue, but I can ask. ;)} AV verification tasks are likely to incur significant CPU hours and therefore are subject to increased exposure to these errors.
\\\\
\noindent\underline{\textit{Game Engine Code:}}
The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh\todo[inline]{introduce actor navigation mesh}, deterministic ego vehicle\todo[inline]{introduce ego vehicle earlier on} controllers and engine texture loading rates especially for perception stack testing. %
For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean?} the total time spent generating a frame, executing the game loop or render thread.
\\\\
\noindent\underline{\textit{Scheduling:}}\todo[inline]{is this at runtime?} %
% info at:
% http://www.informit.com/articles/article.aspx?p=169479&seqNum=3
% https://www.os-book.com/OS8/os8c/slide-dir/PDF-dir/ch5.pdf
% https://en.wikipedia.org/wiki/Concurrency_(computer_science)
% https://en.wikipedia.org/wiki/Concurrent_computing
% https://en.wikipedia.org/wiki/Race_condition#cite_note-1
Scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed dependent on the scheduler policy. A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}. \todo[inline]{is this the scheduling that is done by the operating system, i.e. at runtime?}
If during simulation the scheduling policy was updated or thread priority was inconsistently handled, then a high variance in output may result. It is, however, unlikely that such events would occur on repeated controlled tests for the same hardware and operating system configuration.\todo[inline]{justify - WHY?}
% KIE: If all threads have the same priority, then the scheduler is free to select one, this can give rise to non-det - or not?

Similar to thread scheduling, the CPU scheduling policy determines on which processor core to execute processes in the ready queue and this is maybe ordered by factors such as throughput, latency and CPU utilisation. If the same single-threaded script executes multiple times across different physical cores due to the CPU utilisation policy then should you expect deterministic behaviour? Processor core designs should be identical although impurity inconsistency across the bulk silicon and minor perturbations in the semiconductor processing across the chip may exist. End users expect such minor variations to be controlled for in architecture tolerances and factory testing.\todo[inline]{not clear, what is the difference between scheduling cores and threads - are both done by the operating system?}

@@

\noindent\underline{\textit{Concurrency:}}
% Concurrency is a natural source of nondeterminism since the order and timings of independent processes can vary
\todo[inline]{Say what concurrency is and how executing concurrently running code can lead to non-determinism. Consider also parallelism, e.g. in GPUs. Focus on the order in which the execution occurs, i.e.\ the timing of events, which can cause non-determinism. Say which parts of a game engine could reasonably be expected to run concurrently.}
Another issue to consider is concurrency, where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program. This technique can lead to issues such as shared resource access conflicts and RACE conditions~\cite{huffman1954synthesis}. A game engine may use such techniques to improve performance.\todo[inline]{how?}
% KIE: We only list the potential sources of non-det, I therefore removed this. 
% although monitoring and detecting such events falls outside of the scope for this work.

\noindent\underline{\textit{NUMA Non-Uniform Memory Access:}}
For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the memory location relative to the processor. Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}. 
%
Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically, or, perhaps, simply with an increased data transfer cost. 
%
By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.

% these may. Given the same hardware the output should always be the same, unless scheduling of the threads change. 
% i.e. the utilisation of hardware is manipulated differently each time. This would then result in non-deterministic behaviours especially if it is combined with the floating point problem discussed earlier. 
%
% For example, if a process on thread 1 requires data from thread 2, but thread 2 is delayed due to additional load then synchrony between the threads may be lost if not controlled for explicitly. Given that game engines will promote performance over accuracy the process on thread 1 may be dropped or skipped until the data are available. This may lead to events happening out of order depending on the computational load and hence non-deterministic.
%
% To solve the scheduling of threads problem, one would have to control all of the threads during runtime, which is very involved. To achieve that one would need to replace or control the whole run-time system to allocate tasks to the same threads in the same order. This is beyond the scope of our work and will not be covered in this paper.
\noindent\underline{\textit{Physics and Rendering clocks:}}
\todo[inline]{I'm not sure how VFR impacts on simulation variance, Unity claim that physics will be consistent but 'real-time' is just slowed down.}
% The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic.  A follow up from section \ref{GameLoopSection}, these
Game engines can often have a variable game loop update rate but a fixed physics update rate, e.g.\ Unity\footnote{\url{https://docs.unity3d.com/Manual/TimeFrameManagement.html}}. This is good\todo[inline]{how?} for hardware scalability but creates a challenge for the physics calculations which provide the most consistent results with a small fixed time step. 
%
A variable frame rate\todo[inline]{Is this the game loop update rate?} could contribute to \textit{simulation variance} if the variation in the frame rate is temporally inconsistent between repeated tests and hardware.\todo[inline]{what does the hardware have to do with this? Do we not assume the same HW at this stage?} Given that variable frame rate is designed for optimising performance across differing hardware then inconsistencies across different systems may occur. Repeated tests on the same system should result in similar frame rates unless there is additional competition for resources. 
%
AV specific platforms, such as Carla, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhere to this rate.
%
That is to say, if the frame rate is allowed to vary, then this variation has to be controlled to give the same variability throughout a given test in order for the test results to be repeatable. This issue might be of negligible significance if the frame rate is consistent and the time steps\todo[inline]{of which clock} are small enough.

\noindent\underline{\textit{Actor Navigation:}}
Navigation meshes are used in game engines for actor path planning, and could be considered as a potential source of non-deterministic behaviour, e.g.%. Some AV simulator developers claim that the built-in physics engines' AI is non-deterministic
~\cite{CARLABenchmark}. 
This may depend on the algorithms used by the engine which typically use the A* algorithm~\cite{AStarBook} and should give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}. 
Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
%
It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.\todo[inline]{unclear :(}
% This non-deterministic behaviour can be also resorted to the different scheduling of threads, hence giving variations in meshes every time they are loaded.

@@

>>>>

% The use of game engines for AV testing
Increasingly, the autonomous vehicle community is adopting game engines as simulation platforms to support the development and testing of vehicle control software. 
%
Carla\footnote{http://carla.org/}, for instance, is an open-source simulator for autonomous driving that is implemented as a layer over the Unreal Engine (UE4), a  real-time 3D creation environment for the gaming and film industry as well as other creative sectors~\cite{CARLA_paper}.

@@

% Game engines for SBV
Compared to the vehicle dynamics simulators and traffic-level simulators used by manufacturers~\cite{FrameworkAndChallenges}, game engines offer a convenient simulation solution that meets many of the requirements for the development and functional safety testing of AVs in simulation. However, several challenges remain; one of these is determinism.
%
% What is a deterministic system
Deterministic, in this context, refers to the property of causality given a temporal development of events such that any state is completely determined by prior states.
%
Herein the term \textit{simulation variance} is used to refer to how simulation traces and outputs vary, when the same stimulus is provided repeatedly under identical initial conditions. Specifically, we focus on the variance of actor paths within a given scenario.

We will adopt the terminology defined in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements. Dynamic elements are the elements in a scene whose actions or behaviour may change over time; these are considered actors and may include the AV or \textit{ego vehicle}, other road vehicles, cyclists, pedestrians and traffic signals. The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific parameters.\todo[inline]{values - Does this mean stimulus or other parameters? GC - changed to parameters} A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.

@@

% Challenges of SBV
Game engines do not need to be deterministic when used for gaming; there are no safety implications from non-determinism in this domain, nor is finding and fixing all the bugs a high priority for games developers. Because the situation is very different for AV development and testing, our main research question is: 
how can one assess the extent to which a simulation environment is deterministic? 
%
In pursuit of an answer to this question we have developed a method to determine the simulation variance for a given simulation environment. 
%
Knowing the simulation variance will help developers and testers understand the impact of using a game engine for AV simulation, especially how potentially existing non-determinism affects repeatability and reliability.

\todo[inline]{review and potentially summarize results from case study GC - added see below.}
A case study to highlight the non-deterministic effects in a game engine are presented where scenarios between pedestrian and vehicle actors are used to illustrate conditions that result in high simulation variance. By limiting the the level of system utilisation to less than 75\% and terminating scenarios when any collision callback is made, a deterministic operational domain was found.
%
%The game engine, in this context, must also comply with the requirements to provide a suitably realistic environment and real-time processing for perception and security analysis.\todo[inline]{next}

This paper is structured as follows. In Section~\ref{s:background} game engines and sources of potential non-determinism in game engines are investigated. Section~\ref{s:methodology} presents a method to assess the suitability of a simulation system for AV verification. 
%
A case study to illustrate how the simulation variance observed for a number of scenarios is presented in Section~\ref{s:case-study}.\todo[inline]{point to UE4 GC - ??} We conclude in Section~\ref{s:conclusion} and give an outlook on future work.\todo[inline]{review at end}

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoop.pdf}
\caption{Game engine loop block diagram~\cite{GameProgPatternsBook} \todo[inline]{AG: remove the clock symbol and replace with $dt$, possibly remove the ticks chart as this is confusing given they are not ticks but event markers.}}
\label{GameEngineLoopDiagram}
\end{figure}

The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of AV development, the response of the AV to the current state of its environment. For example, the throttle and steering angle from the controller of an AV. The update stage interprets the input from all dynamic actors in the scene, performs any necessary physics calculations and updates the actor states. The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}.

\todo[inline]{Fig 1: update rates of which clock? GC-is a clock confusing as it does not run on a 'clock' in computer science terms, rather it is a timer which is used as reference to compare to a 'wall clock'}

\todo[inline]{Should there not also be a clock in the overall loop? GC-yes the wall clock but it may get confusing as that is a real clock and the game loop clock is a timer?}

\todo[inline]{Greg to look up Unity wrt concurrently running processes and clocks. Unity page on multi-threading \footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} and a plugin for multi-thread management \footnote{\url{https://forum.unity.com/threads/task-for-unity-scripting-threading.462971/?_ga=2.228561631.331624405.1587372521-1435703724.1587372521}}}

@@

\todo[inline]{We should agree if the 'game clock' really is a clock or just times the computation for each game loop.}

\todo[inline]{For AG: We use the phase 'game clock tick'- where did the come from as I think it is misleading.}

As the game progresses, the game loop runs continuously according to the game clock. \todo[inline]{introduce the different clocks and what they govern, real world, game, physics and render}
%
At the beginning of each game clock tick, the lag\todo[inline]{first explain where the lag comes from} between the game clock and the real world is updated based on how much real time has passed. This measures how far the game's clock is behind compared to the real world.

\todo[inline]{GC maybe replace this entire section with a briefer account: The game loop cycle starts with initialising the scene and actors. Input events from the User or AI are then processed followed by a physics cycle which may repeat more than once per rendered frame if the physics time step, $dt$, is less than the render update rate. The render update will process frames as fast as the computational processing will allow up to the maximum monitor refresh rate\footnote{\url{https://blogs.unity3d.com/2019/06/03/precise-framerates-in-unity/}}. When the frame is rendered the game loop cycle returns to processing inputs.}
%
Then the user inputs are processed.
%
There is then an inner loop to update the physics processes (clock symbol in Fig.~\ref{GameEngineLoopDiagram}), incrementing at a delta time step, $dt$, until the game clock is equal to the real world. 
%
Rendering occurs once the game clock has caught up with the real world clock. 
%
The process then repeats.

The physics engine operates according to a time step, $dt$. %, because it makes everything simpler and more stable for physics and AI.
The shorter this time step is, the smoother the interpretation of the physical dynamics will be. %processing time it takes to catch up to game clock. % and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}. 
%
Fixing the physics time step to a constant $dt$ results in simpler and more stable physics simulation and AI. 
%
To use a fixed physics time step the user's display refresh rate needs to be known in advance. This requires an update loop to take less than one render tick (one frame of real world time). Given the range of different hardware capabilities, a variable delta time is often implemented for game playing, taking the previous frame time as the next $dt$. However, variable $dt$ can lead to different outcomes in repeated tests and in some cases unrealistic physical representations~\cite{gaffer}. 
%
Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allows computational headroom for slower hardware. Some engines provide sub-stepping which processes multiple physics calculations per frame at a greater CPU cost, e.g.\ Unreal Engine\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/}}. 
%
If the engine tries to render between physics updates, \textit{residual lag} can occur and extrapolation between frames needs to be performed to smooth transition between scenes. This is depicted in the lower part of Fig.~\ref{GameEngineLoopDiagram}. \cite{GameEngineArchBook}\cite{GameProgPatternsBook}\todo[inline]{Should this still be cited? GC-No, it is cited elsewhere.}

\todo[inline]{Why is this part of the figure? It would be clearer if it was in a separate figure that shows all the clocks involved and how they interact - depend on each other. GC-suggest we remove as the ticks are not relevant to the main argument or discussion.}
%
% Ideally the time step should be as short as possible, so that the simulation returns better accuracy of complex physical processes. %runs with high fidelity on fast machines. 
In exceptional cases where resources are scarce, the fixed time step can be greater than the time between render ticks %it takes to process an ``Update" inner loop on some (slow) machines, 
and the simulation will exhibit lag between input commands and rendered states, resulting in unsynchronised and unrealistic behaviour.

\todo[inline]{Wrt the figure, is ``lag'' not a delay, i.e.\ should happen after the tick, rather than before? GC-suggest remove this part of the diagram, too confusing as there are not really any clocks or ticks - we should remove all mention of ticks too.}

@@

In the context of AV simulation, this could result in precision issues of, for example, actor position within the environment leading to falsely satisfied or failed assertions.\todo[inline]{state what assertions are (used for) GC-see below} In the field of autonomous vehicle verification, assertions are the immutable rules and regulations that describe safe and legal driving. 

@@

%Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.increasingl\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}. 

@@

\noindent\underline{\textit{Error Correcting Code (ECC) Memory:}}
ECC Memory can detect and correct single bit errors in DRAM memory. Single bit errors may occur due to malfunctioning hardware, ionising radiation (background cosmic or environmental sources) or from electromagnetic radiation~\cite{dodd2003basic}. If single bit errors go uncorrected then subsequent computational processing will produce incorrect results, potentially giving rise to non-determinism due to the probabilistic nature of such errors occurring. Estimating the rate of error is difficult and dependent on hardware, environment and computer cycles~\cite{mielke2008bit}.\todo[inline]{that said, HW simulation is deterministic, uses long simulation runs and thousands of CPU hours - I've not heard that this is an issue, but I can ask. ;) GC-yes although most commercial systems and servers will use ECC, although some development SMEs may use non-ECC and therefore should be aware of this issue.} AV verification tasks are likely to incur significant CPU hours and therefore are subject to increased exposure to these errors.

\noindent\underline{\textit{Actor Navigation:}}
There is existing evidence to suggests actor navigation, typically pedestrians, could be the cause of non-deterministic simulation behaviour~\cite{CARLABenchmark}.
% 
Unreal Engine is the underlying framework for the Carla simulator which uses the A* algorithm for actor navigation\footnote{\url{https://www.oreilly.com/library/view/unreal-engine-4/9781784393120/ch04s03.html} from~\cite{newton2016unreal}}.
%
The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}. 
%
Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in the environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity.%. Some AV simulator developers claim that the built-in physics engines' AI is non-deterministic
%
However, the environment management in Carla implements fixed binary files for navigation meshes that are linked to each driving scene and cannot be unknowingly modified and therefore can be considered fixed. 
%
Unless other sources that can alter the fixed environment exist, actor navigation should not be considered a source of non-determinism
%
The navigation mesh is a fixed area of the environment where actors are free to navigate. Typically the navigation mesh is fixed for a specific environment or road scene and therefore the A* algorithm
 should give deterministic results.

% Therefore, the determinism of the actor path planning depends on the environment and the update to actor states within it and how navigation meshes are created or modelled, e.g. shape, granularity. 
%
% It is interesting to note that changes could occur to navigation meshes if they are dynamically loaded as the simulation proceeds which can reduce computational load for large area maps\footnote{\url{https://docs.unrealengine.com/en-US/BlueprintAPI/AI/Navigation/RegisterNavigationInvoker/}}.
\todo[inline]{unclear :( GC- re-written}

\noindent\underline{\textit{Game Engine Code:}}
The type and version of the engine code executed should be considered, paying attention to the control of pseudo-random numbers, fixed physics calculation steps ($dt$), fixed actor navigation mesh\todo[inline]{introduce actor navigation mesh GC-section on actor navigation brought forward}, deterministic ego vehicle\todo[inline]{introduce ego vehicle earlier on GC added line 99} controllers and engine texture loading rates especially for perception stack testing. %
For example, in Unreal Editor the \textit{unit}\footnote{\url{https://docs.unrealengine.com/en-US/Engine/Performance/StatCommands/}} command can be used to determine processing bottlenecks by indicating\todo[inline]{what does indicating mean? GC-displaying, returning or showing} the total time spent generating a frame, executing the game loop or render thread.
\\\\
\noindent\underline{\textit{Scheduling:}}\todo[inline]{is this at runtime?} %
% info at:
% http://www.informit.com/articles/article.aspx?p=169479&seqNum=3
% https://www.os-book.com/OS8/os8c/slide-dir/PDF-dir/ch5.pdf
% https://en.wikipedia.org/wiki/Concurrency_(computer_science)
% https://en.wikipedia.org/wiki/Concurrent_computing
% https://en.wikipedia.org/wiki/Race_condition#cite_note-1
Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed dependent on the scheduler policy. A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}. \todo[inline]{is this the scheduling that is done by the operating system, i.e. at runtime? GC - runtime, added to 1st para.}
%
If during simulation the scheduling policy was updated or thread priority was inconsistently handled, then a high variance in output may result. It is, however, unlikely that such events would occur on repeated controlled tests for the same hardware and operating system configuration.\todo[inline]{justify - WHY? GC- see below} Given that hardware is considered deterministic and if the scheduling policy does not change between tests, then for the same hardware and operating system configuration the same output will be given. 
%
However, if some aspects of the game loop are multi-threaded\footnote{\url{https://docs.unity3d.com/Manual/JobSystemMultithreading.html}} and the scheduling policy is set to equal priority for all threads then a background process may interrupt the otherwise deterministic sequence of events which may, for example, alter the number of physics calculations that can be performed within the game loop and hence variance in the simulation. 
% KIE: If all threads have the same priority, then the scheduler is free to select one, this can give rise to non-det - or not?

Similar to thread scheduling, the CPU scheduling policy determines on which processor core to execute processes in the ready queue and this is maybe ordered by factors such as throughput, latency and CPU utilisation. If the same single-threaded script executes multiple times across different physical cores due to the CPU utilisation policy then should you expect deterministic behaviour? Processor core designs should be identical although impurity inconsistency across the bulk silicon and minor perturbations in the semiconductor processing across the chip may exist. End users expect such minor variations to be controlled for in architecture tolerances and factory testing.\todo[inline]{not clear, what is the difference between scheduling cores and threads - are both done by the operating system? GC - does this come from a software request? i.e. request to use a core or just a thread}

@@

\noindent\underline{\textit{Concurrency:}}
% Concurrency is a natural source of nondeterminism since the order and timings of independent processes can vary
\todo[inline]{Say what concurrency is and how executing concurrently running code can lead to non-determinism. Consider also parallelism, e.g. in GPUs. Focus on the order in which the execution occurs, i.e.\ the timing of events, which can cause non-determinism. Say which parts of a game engine could reasonably be expected to run concurrently.}

Concurrency is where sequential tasks can be executed out of order to improve speed by utilising parallelisation depending on the decomposability of the program and may be another source of non-determinism to consider. This technique can lead to issues such as shared resource access conflicts and RACE conditions~\cite{huffman1954synthesis}. A game engine may use techniques such as multi-threading or GPU parallelisation to improve performance by offloading time-critical calculations to several resources simultaneously which would be quicker than a serial execution.\todo[inline]{how? GC-see above.}
% KIE: We only list the potential sources of non-det, I therefore removed this. 
% although monitoring and detecting such events falls outside of the scope for this work.

\noindent\underline{\textit{NUMA Non-Uniform Memory Access:}}
For a repeated test that operates over a number of cores based on a CPU scheduling policy, memory access time may vary depending on the physical memory location relative to the processor. Typically a core can access its own memory with lower latency than that of another core resulting in lower interprocessor data transfer cost~\cite{nieplocha1996global}. 
%
Changes in latency between repeated tests may, in the worst case, cause the game engine to operate non-deterministically, or, perhaps, simply with an increased data transfer cost. 
%
By binding a process to a specific core for the duration of its execution, the variations in data transfer time can be minimised.

% these may. Given the same hardware the output should always be the same, unless scheduling of the threads change. 
% i.e. the utilisation of hardware is manipulated differently each time. This would then result in non-deterministic behaviours especially if it is combined with the floating point problem discussed earlier. 
%
% For example, if a process on thread 1 requires data from thread 2, but thread 2 is delayed due to additional load then synchrony between the threads may be lost if not controlled for explicitly. Given that game engines will promote performance over accuracy the process on thread 1 may be dropped or skipped until the data are available. This may lead to events happening out of order depending on the computational load and hence non-deterministic.
%
% To solve the scheduling of threads problem, one would have to control all of the threads during runtime, which is very involved. To achieve that one would need to replace or control the whole run-time system to allocate tasks to the same threads in the same order. This is beyond the scope of our work and will not be covered in this paper.
\noindent\underline{\textit{Physics and Rendering clocks:}}
\todo[inline]{I'm not sure how VFR impacts on simulation variance, Unity claim that physics will be consistent but 'real-time' is just slowed down.}
% The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic.  A follow up from section \ref{GameLoopSection}, these
Game engines can often have a variable game loop update rate but a fixed physics update rate, e.g.\ Unity\footnote{\url{https://docs.unity3d.com/Manual/TimeFrameManagement.html}}. This is good\todo[inline]{how?} for hardware scalability but creates a challenge for the physics calculations which provide the most consistent results with a small fixed time step. 
%
A variable frame rate\todo[inline]{Is this the game loop update rate?} could contribute to \textit{simulation variance} if the variation in the frame rate is temporally inconsistent between repeated tests and hardware.\todo[inline]{what does the hardware have to do with this? Do we not assume the same HW at this stage? GC- we are just listing sources at this point so should be hardware agnostic. also, yes, hardware is the same for us but we cannot control for resource utilisation and background processes.} Given that variable frame rate is designed for optimising performance across differing hardware then inconsistencies across different systems may occur. Repeated tests on the same system should result in similar frame rates unless there is additional competition for resources. 
%
AV specific platforms, such as Carla, can be operated with fixed frame rates which should provide maximum stability as long as the frame rate during execution (FPS) is monitored and observed to adhere to this rate.

@@

\noindent\underline{\textit{Summary:}}
Floating point as long as h/w is det any deterministically-incorrect

ECC: unless you live in radiation heavy area or doing significant repeated tests unlikely to see errors unless there is specific hardware faults - can be checked with memory testing software

Actor navigation not a problem if navigation meshes are fixed

Game engine, ensuring fixed $dt$ and control of random numners

Core or thread scheduling will not be issue unless changing scheduling policy or migrating bewtween machines that have different policies.

Concurrency is really down to the development of the simulation software and reputable companies should have good control over these potential issues following good software development practice.

NUMA should only affect interprocessor data transfer cost, at wrst only make computations longer. relative times beyween access different caches is likely to be small although may impact on high throughput, e.g. HPC.

VFR - want to remove this section.
